A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-18 analysis/paper/paper.Rmd
edit the main text in paper rmd

<!---network analysis approaches and basic concepts-->
Network analysis has been increasingly used by archaeologists to understand past relationships, interactions, and structures of observed phenomena by visualizing and analyzing rational data [@Brandes2013; @Mills2017; @Peeples2019]. The recent development of modeling approaches enable hypothesis testing for the intercorrelations between individual elements and overall structures inherent in networks [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. By comparing hypothesized networks with the observed network, network modeling can answer anthropological questions related to exchange [@Gjesfjeld2013; @Crabtree2015], diffusion [@Ostborn2015], or social transformation [@Mills2013] in a statistical way. Among network modeling methods, exponential random graph models (ERGMs) are stochastic models for investigating the process of network formation through dependence assumptions for relationships and simulations for network patterns [@Harris2013; @Ghafouri2020]. Since relationships in a network emerging relative to their position in a network can be hypothesized, ERGMs are promising to evaluate dynamic social processes behind observed archaeological networks [@Brughmans2014]. However, computational difficulties and the sensitivity to uncertainties to some extent limits the practical applications of ERGMs, where results could be challenging for interpretation [@Caimo2014].

@@

<!--theoretical framework and research questions-->
The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000], while corporate-based societies stress communal ritual elements, shared power, and wealth differentiation, if any, would be associated with corporate groups, such as age groups [@Siegel1999]. Using the theoretical framework of corporate-network strategies, we test a hypothesis that changes from a corporate mode (a structure showing more subgroups with less wealth differentiation ) to a network mode (a centralized structure with more wealth differentiation) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until arrival of Han Chinese immigrants).

@@

<!-- Our chronology and principles of network building -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction because we cannot determine their chronology. Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. These items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as having higher status in their community [@LiandWu2006].

@@

## Hypothesis and construction of networks {-}
<!-- our expectation based on hypothesis --> 
Our pericolonial impact hypothesis is that differential access to foreign prestige goods after the European presence led to an increased social inequality at Kiwulan, where social structure changed from a more corporate strategy to a more network strategy [@Blanton1996; @Drennan2010; @Feinman2000]. This would be reflected by the pattern of intercorrelations of foreign prestige goods in burial networks. If social inequality gradually increased as we hypothesize, then we expect to observe a network structure with higher centralization (or popularity) and less transitivity (or clustering), along with associations with wealth accumulation, indicative of a network-based society. In contrast, the network before the European arrival will show low centralization and high transitivity, associated with ritual element, or age/sex groups to indicate a corporate-based society. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they both possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and a burial value index. Ritual pottery was identified as locally made ceramics placed above an individual. Those pots mostly have soot on vessel exteriors and charred residues on interiors that are interpreted as vessels for funeral feasting [@Hsieh2009]. We calculated our burial value by summing values of all types of prestige goods in each burial context. The value of prestige good is calculated by taking the total number of burials in Kiwulan, and then dividing this by the number of burials with each type of prestige good [@Jorgensen1991]. We then assigned the burials into three ranks according to their burial values, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth (Details about other node attributes in Supplementary Online Materials). Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

@@

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
ergm_terms_selected <-
  ergm_terms %>% 
  select(-"Description", -"note")
knitr::kable(ergm_terms_selected,
             caption = "Network parameters in ERGMs used for model specifications with associated archaeological interpretation for burial relations.")
```

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models, which follows a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))). For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. For physical distance between burials, we also set a vague prior effect (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological model of increased social inequality over time based on assumptions of corporate-network strategies [@Drennan2010; @Feinman2000], we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity (gwesp) and centralization (gwedegree). We set the priors to higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)) for the network before European contact to indicate less social differentiation and stress ritual element shared in corporate groups. Conversely, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and higher covariate effect of burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)) to model an increased social differentiation.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). A significant effect is determined by consistent positive/negative estimates in a 95% confidence interval. For nodal covariates in the pre-European model, ritual element represent by pots (ritual-homophily) has a significant effect on the formation of relations between burials, while wealth rank represented by burial values (wealth-homophily) has a negative effect. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same wealth level tend not to. Despite positive mean for some covariates, such as age (age-homophily) and sex (sex-homophily), they do not show a significant tendency due to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not reflect similarity in burial treatment. For the endogenous network effects, transitivity (gwesp) presents a significant positive effect, while centralization (gwdegree) demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials have a similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter= case_when(parameter == "edges" ~ "Density",
                              parameter == "nodematch.age" ~ "Age-homophily",
                              parameter == "nodematch.sex" ~ "Sex-homophily",
                              parameter == "nodematch.ritual" ~ "Ritual-homophily",
                              parameter == "nodematch.value" ~ "Wealth-homophily",
                              parameter == "gwesp" ~ "tansitivity",
                              parameter == "gwdeg" ~ "centralization",
                              parameter == "dyadcov.distance" ~ "Physical distance")) %>% 
  mutate(parameter = factor(parameter,
                            levels=c("Density", "Age-homophily", 
                                     "Sex-homophily", "Ritual-homophily",
                                     "Wealth-homophily", "tansitivity",	
                                     "centralization", "Physical distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.975), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.3) %>% 
  width(j = 2:9, width = 0.65) %>% 
  set_caption("Estimated posterior means, medians, and 95% confidence intervals for each network parameter of two models. Confidence intervals that do not include zero idicate significant effects of parameters (Caimo, 2017).")
ft
```

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, gwesp, gwdegree, and distance-dyadcov by phases. The pre-European group presents remarkable larger values for the gwesp, but lower values for the edges and gwdegree parameters compared to the post-European group. The distance parameters overlap between two groups."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
For the network model after the arrival of Europeans, the nodal covariates of wealth rank show significant positive effects, while ritual and sex variables have negative effects. This indicates the burials in the same wealth level tend to have relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to have relations. There are no significant effects for variables of age and physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the centralization (gwdegree) shows a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation or display in burial events. In general, the post-European network model has a smaller transitivity effect and a positive centralization effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. In addition, both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used vertex bootstrap to cross-validate the results of Bayesian ERGMs. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) that is useful for quantifying standard errors and estimating sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks by examining their confidence intervals for the network population. Networks with nonoverlapped intervals means there is a significant difference. We computed and compared endogenous network statistics, including density, centralization, and transitivity for our two networks. Figure \@ref(fig:figure-bootstrap-CI) shows a significant difference in observed network centralization between the two networks, which is consistent with our finding of negative centralization in the pre-European period and positive centralization in the post-European period using the Bayesian approach. For density and transitivity, the 95% confidence intervals from the two networks overlapped with each other that indicates no significant difference. This is also consistent with our results of Bayesian modeling, where both networks present similar positive or negative effects with some degree of difference.

```{r figure-bootstrap-CI, fig.cap = "The 95% bootstrap confidence intervals for the observed networks, the top one with 0 nodes removal, and networks with different numbers of node removal. The number of bootstrap samples is 1000."}
knitr::include_graphics(here::here("analysis", "figures", "007-bootstrap-CI.png"))
```

```{r bootstrap-t-test, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-bootstrap-t-test.R"))}
suppressWarnings(source(here::here("analysis", "paper", "008-bootstrap-t-test.R")))
pval_den <- signif(density_b_test$p.value, 4)
pval_trans <- signif(trans_b_test$p.value, 4)
pval_star <- signif(star_b_test$p.value, 4)
```

The vertex bootstrap procedure with a t-test on the difference of network populations also shows that there are significant differences in network density(p = $`r pval_den`$), transitivity (p = $`r pval_trans`$), and centralization (p = $`r pval_star`$). Moreover, we explored the sample size effect by removing nodes at certain percent (5-40%) for both networks and compared their confidence intervals. The results (Figure \@ref(fig:figure-bootstrap-CI)) demonstrate a consistent difference between the two networks within 30 % node removals that suggest they are robust to the different sizes. In general, those statistical methods support the presence of anthropologically meaningful differences for density, transitivity, and centralization (i.e. centrality) in the pre-and post-European networks. The procedure of cross-validation ensures our results can provide robust comparison between two networks without sample size effects.

>>>>

<!---network analysis approaches and basic concepts-->
Network analysis has been increasingly used by archaeologists to understand past relationships, interactions, and structures of observed phenomena by visualizing and analyzing rational data [@Brandes2013; @Mills2017; @Peeples2019]. The recent development of modeling approaches enable hypothesis testing for the intercorrelations between individual elements and overall structures inherent in networks [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. By comparing hypothesized networks with the observed network, network modeling can answer anthropological questions related to exchange [@Gjesfjeld2013; @Crabtree2015], diffusion [@Ostborn2015], or social transformation [@Mills2013] in a statistical way. Among network modeling methods, exponential random graph models (ERGMs) are stochastic models for investigating the process of network formation through dependence assumptions for relationships and simulations for network patterns [@Harris2013; @Handcock2008; @Ghafouri2020]. Since relationships in a network emerging relative to their position in a network can be hypothesized, ERGMs are promising to evaluate dynamic social processes behind observed archaeological networks [@Brughmans2014]. However, computational difficulties and the sensitivity to uncertainties to some extent limits the practical applications of ERGMs, where results could be challenging for interpretation [@Caimo2014].

@@

<!--theoretical framework and research questions-->
The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000], while corporate-based societies stress communal ritual elements, shared power, and wealth differentiation, if any, would be associated with corporate groups, such as age groups [@Siegel1999]. Using the theoretical framework of corporate-network strategies, or horizontal hierarchy, we test a hypothesis that changes from a corporate mode (a structure showing more subgroups with less wealth differentiation ) to a network mode (a centralized structure with more wealth differentiation) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). The preservation of human remains at Kiwulan was generally poor, so we have incomplete age and sex data (details in Supplementary Online Materials) and lack health status data. Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until arrival of Han Chinese immigrants).

@@

<!-- Our chronology and principles of network building -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction because we cannot determine their chronology.

Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. These items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as having higher status in their community [@LiandWu2006].

@@

## Hypothesis and construction of networks {-}
<!-- our expectation based on hypothesis --> 
Our pericolonial impact hypothesis is that differential access to foreign prestige goods after the European presence led to increased social inequality at Kiwulan, where social structure changed from a more corporate strategy to a more network strategy [@Blanton1996; @Drennan2010; @Feinman2000]. This would be reflected by the pattern of intercorrelations of foreign prestige goods in burial networks. If social inequality gradually increased as we hypothesize, then we expect to observe a network structure with higher centralization (or popularity) and less transitivity (or clustering), along with associations with wealth accumulation, indicative of a network-based society. In contrast, the network before the European arrival will show low centralization and high transitivity, associated with ritual element, or age/sex groups to indicate a corporate-based society. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they both possess each type of goods (i.e. presence or absence). Node attributes here include osteologically determined age and sex, cultural data, such as ritual pottery, and a burial value index. Ritual pottery was identified as locally made ceramics placed above an individual. Those pots mostly have soot on vessel exteriors and charred residues on interiors that are interpreted as vessels for funeral feasting [@Hsieh2009]. We calculated our burial value by summing values of all types of prestige goods in each burial context. For example, for gold-foil beads, we take the total number of burials (90) and divide by number of burials with gold-foil beads (46), to give a value of 90/46 = 1.96. We repeat this for each type of prestige item. For each burial we sum these values of each type of prestige item to get a burial value [@Jorgensen1991].

We then assigned the burials into three ranks according to breaks in the distribution of burial values, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth (details in Supplementary Online Materials). Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

@@

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
ergm_terms_selected <-
  ergm_terms %>% 
  select(-"Description", -"note")
knitr::kable(ergm_terms_selected,
             caption = "Network parameters in ERGMs used for model specifications with associated archaeological interpretation for burial relations. ERGM terminology is from the R package statnet (Handcock, 2008; Morris, 2008)")
```

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models, which follows a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))). For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. For physical distance between burials, we also set a vague prior effect (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological model of increased social inequality over time based on assumptions of corporate-network strategies [@Drennan2010; @Feinman2000], we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity and centralization. We set the priors to higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)) for the network before European contact to indicate less social differentiation and stress ritual element shared in corporate groups. Conversely, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and higher covariate effect of burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)) to model an increased social differentiation.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the ritual element represent by pots (ritual-homophily) has a significant effect on the formation of relations between burials. On the other hand, the wealth rank, represented by burial values (wealth-homophily), has a negative effect. This demonstrates that burials in the pre-European period with ritual pottery tend to form relations, but burials in the same wealth level tend not to. Despite positive mean values for some covariates, such as age and sex, their confidence intervals include zero, so they do not show a significant influence. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties, without any prominent burials. This might imply that individuals have equal access to the flow of trade goods.

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter= case_when(parameter == "edges" ~ "Density",
                              parameter == "nodematch.age" ~ "Age-homophily",
                              parameter == "nodematch.sex" ~ "Sex-homophily",
                              parameter == "nodematch.ritual" ~ "Ritual-homophily",
                              parameter == "nodematch.value" ~ "Wealth-homophily",
                              parameter == "gwesp" ~ "transitivity",
                              parameter == "gwdeg" ~ "centralization",
                              parameter == "dyadcov.distance" ~ "Physical distance")) %>% 
  mutate(parameter = factor(parameter,
                            levels=c("Density", "Age-homophily", 
                                     "Sex-homophily", "Ritual-homophily",
                                     "Wealth-homophily", "transitivity",	
                                     "centralization", "Physical distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.975), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.3) %>% 
  width(j = 2:9, width = 0.65) %>% 
  set_caption("Estimated posterior means, medians, and 95% confidence intervals for each network parameter of two models. Confidence intervals that do not include zero indicate significant effects of parameters (Caimo, 2017). Posterior means are generally close to posterior medians.")
ft
```

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, gwesp, gwdegree, and distance-dyadcov by phases. The pre-European group presents remarkable larger values for the gwesp, but lower values for the edges and gwdegree parameters compared to the post-European group. The distance parameters overlap between two groups. Both posterior estimates present generally symmetric distributions."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
After the arrival of Europeans, the nodal covariates of wealth rank show significant positive effects, while the ritual and sex variables have negative effects. This indicates that burials in the same wealth level tend to be more connected with each other. On the contrary, the same sex burials, and burials with ritual pottery in common tend, not to have relations. In the post-European burials there are no significant effects for variables of age and physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but much weaker than in the pre-European network. In contrast, centralization has a significantly higher positive effect than the pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials having many more ties than others. This implies more access to trade goods and the increased wealth accumulation and display behaviors in burial events. In general, the post-European network model has a smaller transitivity effect and a positive centralization effect compared to the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 30% node removals (Figure \@ref(fig:figure-bootstrap-CI)). This indicates that the network variables under investigation are robust to the different network size of our sample. Figure \@ref(fig:figure-bootstrap-CI also shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity.

```{r figure-bootstrap-CI, fig.cap = "Results of the vertex bootstrap analysis by three network variables, showing the 95% bootstrap confidence intervals for the bootstrapped (n = 1000) networks, for each category of node removal. The 0% node removal category is the original data for the two phases. Networks with nonoverlapping confidence intervals indicate a significant difference between the two phases for that network variables."}
knitr::include_graphics(here::here("analysis", "figures", "007-bootstrap-CI.png"))
```
